Error-Centric Annotation of Learner Corpora

### 2009-04-09 text/paper.tex
bbbbbbbbbbbbbbbbbbbb

\documentclass[12pt]{article}
\usepackage[utf8x]{inputenc}
\usepackage{graphicx,linguex,natbib,aleks,amsmath,amsthm}
\author{Magdalena Leshtanska \and Aleksandar Dimitrov}
\title{Error-Centric Annotation of Learner Corpora}
\begin{document}
\maketitle

@@

Recently, there has been a growing interest in learner language corpora and
learner errors analysis (citation). There is a need to
identify errors in text, to make the automatic processing of those texts
feasible\cite{nessel}. Also, the potential of learner errors for educational purposes and language
acquisition insight has become increasingly recognized (citation), hence, many
annotation schemes have been devised to identify and classify learner errors,
such as \cite{negthesis}

We propose a novel approach to annotating learner errors, based on the idea that
an error is an inconsistency with its surroundings, rather .  This basic markup approach can be succesfully
combined with any taxonomy for error ??classification. We have coupled it with a
taxonomy for error classification and used the resulting scheme to annotate a
subset of the NOce corpus (citation), achieving an inter-annotation agreement of
???percent.

Firstly, every annotation scheme that operates directly on the text or spans
over a piece of text will run into two kinds of problems:

Based on these assumptions, we decided to decouple the \textit{error markup}
from the \textit{corpus data}. Specifically, to our annotation method, the
errors and the text are two entirely different data structures. Every
particular error can reference tokens within the corpus using a \textbf{key},
similar to the way modern Relational Data Bases reference their data. This makes
the index more accessible and easier to maintain and eliminates both problems
above, because every single annotation can be completely independent from all
other annotations.

@@

In order to identify errors occurring in learner language, one has to first
classify them. However, even then, identifying which class a particular error
belongs to is in no way an easy job. Devising an exhaustive taxonomy of errors
that can appear in natural language seems a daunting task, since science has so
far failed to regularize what \textit{is} a valid utterance of a language.
Therefore, our taxonomy will try to do some things only, and do them well. In
particular, we will not cover punctuation mistakes, since their possible
corrections can have cascading effects on the rest of the sentences and suddenly
give rise to all kinds of new errors.

@@

For our test taxonomy, we used the taxonomy described in \cite{negrillo} as a
base, and enhanced it by several concepts from the CLC annotation scheme, as
specified in \cite{CLC}. While we retained the general ideas and structure of
the aforementioned schemata, we aimed to improve their generality, and
systematization.

@@

\begin{itemize}
  \item \textit{Normative Errors.} Some normative concepts are present in almost
  all European languages, such as capitalization at the beginning of a sentence.
  We chose to ignore such errors, since they are not a sign of weakness in the
  target language but likely pure by chance mistakes.
  \item \textit{Style Errors.} Style is too soft a concept. Errors in style are
  notoriously hard to peg and reliably quantify. Moreover, the NOCE corpus
  contained data mostly from beginning English learners, where style errors are
  not as important, evident or relevant.
  \item \textit{Punctuation.} Punctuation mistakes are a very delicate concern:
  they can have tremendous effects on any given token string on both a syntactic
  and semantic level. They might change constituent boundaries, sentence
  boundaries, argument structures and many more things. Moreover, they tend to
  have cascading effects on the correctness of a given sentence. Again, because
  we were annotating a beginner's corpus, we chose to ignore these kinds of
  errors, since in beginner language punctuation often degrades to line noise
  very quickly.
\end{itemize}

@@

The source files for our corpus consisted of plain text files from the
NOCE-corpus % FIXME cite
of beginning to intermediate Spanish learners of English. Our corpus
format is defined in \ref{def:errorformat}

@@

\subsection{Limitations}
Some elements, for example word order errors have enforced context, but it is hard to express via our schema, so we don't do it. 
No correct positions specified by word order errors - also hard to express. One could assign them a target position after an existing token, but then one must also keep track of other "reorderings" by multiple word order errors - I would like beer some to drink - here, both erroneous units - some and to drink should be placed between like and beer, but one must take care to order them correctly. Also, as our schema tags all possible errors, in the case of interchangeable errors (cf XXX), one cannot be certain if the error annotated is really the one, made by the learner, and not the other possible one. Most  importantly, there are sometimes more than one places where a word sequence can fit in.

Errors within a token cannot be analyzed – for example, agglutination, because a token is our smallest unit.

@@

Statistical inter-annotator agreement  measures are a common quality assessment
method used in corpus linguistics and related fields. Hereby, annotations made
on a particular data set by two or more annotators are compared using
quantitative methods.  \cite{ap2008} give an overview of currently employed
methods.

@@

\cite{ap2008} briefly discuss unitization and go on to note that it has thus far
not been exhaustively researched. Even more importantly, they explicitly comment
on the unknown status of the validity of the only inter-annotator agreement
measure in the corpus linguistic literature, $\alpha_U$, presented in
\cite{krip1995}.  Apart from being untested, $\alpha_U$ also seems to have
problems with overlapping markup both in the text, and between annotation sets
(both of which are frequent in learner language data.) The measure also assumes
annotation spans to be continuous, which is not the case in our data.

@@

After assessing the quality of our data, we reached
the conclusion that the error format described in \ref{sec:errorformat}
might benefit from several refinements. Adding a field for part of speech tags
might contribute to the clarity of the data, as well as to its searchability.
The annotated corpus could be queried for erroneously placed verbs or
prepositions, for example.

@@

A part of speech tagger could aid the annotators and enhance the corpus data
significantly. However, there are only a few reports on reliable part of speech
tagging for learner language. \cite{luedeling} presents one such approach.

@@

\subsection{Compound nouns:}

@@

@@

\section{Source Code and XML Data}\label{sec:github}

@@

>>>>

\documentclass[12pt]{article}
\usepackage[utf8x]{inputenc}
\usepackage{graphicx,linguex,natbib,aleks,amsmath,amsthm}
\bibpunct{(}{)}{;}{a}{,}{,}
\bibliographystyle{plainnat}
\author{Magdalena Leshtanska \and Aleksandar Dimitrov}
\title{Error-Centric Annotation of Learner Corpora}
\begin{document}
\maketitle

@@

Recently, there has been a growing interest in learner language corpora and
learner errors analysis \cite{nessel}. There is a need to identify errors in
text produced by second language learners, to make the automatic processing of
those texts feasible (ibid.). Also, the potential of learner errors for
educational purposes and language acquisition insight has become increasingly
recognized (ibid.), hence, many annotation schemes have been devised to
identify and classify learner errors, such as \cite{negrillo}, \cite{CLC},
\cite{ludeling}

We propose a novel approach to annotating learner errors, based on the idea that
an error is an inconsistency with its surroundings. This basic markup approach
can be  combined with a taxonomy for error classification. We
have coupled it with learner error  classification system and used the resulting
scheme to annotate a subset of the NOCE corpus (\cite{negrilloph}), a corpus of
beginning to intermediate English learners of Spanish.

Section \ref{sec:scheme} will introduce the general idea behind our annotation
scheme and elaborate on more technical details. Section 3  gives a brief
overview of real world performance of our method, as far as this is indeed
possible. An annotation manual is contained in Appendix A.

@@

Devising an exhaustive taxonomy of errors that can appear in natural language
seems a daunting task, since science has so far failed to regularize what
\textit{is} a valid utterance of a language.  Therefore, our annotation scheme
strives  to  do some things only, and do them well.

@@

For our test taxonomy, we used the taxonomy described in \cite{negrilloph} as a
base, and enhanced it by several concepts from the CLC annotation scheme, as
specified in \cite{CLC}. While we retained the general ideas and structure of
the aforementioned schemata, we aimed to improve their generality, and
systematization.

@@

\begin{itemize}
  \item \textit{Normative Errors.} Some normative concepts are present in almost
  all European languages, such as capitalization at the beginning of a sentence.
  We chose to ignore such errors, since they are not a sign of weakness in the
  target language but likely pure by chance mistakes.
  \item \textit{Style Errors.} Style is too soft a concept. Errors in style are
  notoriously hard to peg and reliably quantify. Moreover, our corpus
  contained data mostly from beginning English learners, where style errors are
  not as important, evident or relevant.
  \item \textit{Punctuation.} Punctuation mistakes are a very delicate concern:
  they can have tremendous effects on any given token string on both a syntactic
  and semantic level. They might change constituent boundaries, sentence
  boundaries, argument structures and many more things. Moreover, they tend to
  have cascading effects on the correctness of a given sentence. Again, because
  we were annotating a beginner's corpus, we chose to ignore these kinds of
  errors, since in beginner language punctuation often degrades to line noise
  very quickly.
\end{itemize}

@@

The context mechanism in our annotation scheme can increase the complexity of the annotation
process drastically. Since our annotation technique requires us to build a
non-hierarchical graph structure over our input\footnotemark, a traditional
sequential markup scheme scheme that operates directly on the text or spans
over a piece of text will run into at least two kinds of problems:

\footnotetext{Relating words and their dependencies more like a net than as a
hierarchically nested span over a character sequence}

Based on these assumptions, we decided to decouple the \textit{error markup}
from the \textit{corpus data}. Specifically, to our annotation method, the
errors and the text are two entirely different data structures. Every particular
error can reference tokens within the corpus using a \textbf{key}, similar to
the way modern Relational Data Bases reference their data. This makes the index
more accessible and easier to maintain and eliminates both problems above,
because every single annotation can be completely independent from all other
annotations.  The source files for our corpus consisted of plain text files from
the NOCE-corpus of beginning to intermediate Spanish learners of English. Our
corpus format is defined in \ref{def:errorformat}.

@@

\subsection{Limitations}
Word order errors still pose a particular complication even to a non-sequential
scheme such as ours. We are not able to account for the complex dependencies
occurring in some word order mistakes simply by a dependency-graph.

Since our minimal atomic unit is a token, we depend on a tokenization process.
Furthermore, we cannot account for mistakes that occur on a scale smaller than a
token.

@@

Statistical inter-annotator agreement  measures are a common quality assessment
method used in corpus linguistics and related fields. Hereby, annotations made
on a particular data set by two or more annotators are compared using
quantitative methods.  \cite{artstein} give an overview of currently employed
methods.

@@

\cite{artstein} briefly discuss unitization and go on to note that it has thus far
not been exhaustively researched. Even more importantly, they explicitly comment
on the unknown status of the validity of the only inter-annotator agreement
measure in the corpus linguistic literature, $\alpha_U$, presented in
\cite{krip}.  Apart from being untested, $\alpha_U$ also seems to have
problems with overlapping markup both in the text, and between annotation sets
(both of which are frequent in learner language data.) The measure also assumes
annotation spans to be continuous, which is not the case in our data.

@@

After assessing the quality of our data, we reached
the conclusion that the error format described in \ref{sec:taxonomy}
might benefit from several refinements. Adding a field for part of speech tags
might contribute to the clarity of the data, as well as to its searchability.
The annotated corpus could be queried for erroneously placed verbs or
prepositions, for example.

@@

A part of speech tagger could aid the annotators and enhance the corpus data
significantly. However, there are only a few reports on reliable part of speech
tagging for learner language. \cite{ludeling} presents one such approach.

@@

\subsection{Compound nouns}

@@

\subsection{Sequences}
For both triggers and erroneous tokens, it is best to use a single token. In
case of a phrase, only take the keyword and at most any relevant information
bearing token. In such a case, do not care about order.

Still, fixed expressions, such as  Collocations or Idiom can be counted as one
unit. In such a case, annotate the whole phrase, even if it contains erroneous
tokens. Only do not annotate the target if it is within the idiom.

@@

\section{Source Code and XML Data}\label{appendix:github}

@@

\bibliography{bib}
